---
title: "R Refresher -- Part 4: Applied Case Study (Chart Selection)"
output:
  html_notebook:
    toc: true
    toc_depth: 2
    number_sections: false
---

=======================;=======================================================
R PROGRAMMING REFRESHER -- PART 4: APPLIED CASE STUDY
CHART SELECTION FOR school_district.R / school_district_v2.R
Continues the numbering from Part 3 (which ended at Section 35). Answers a
real question directly: given the actual variables in the school-district
scripts, which chart fits, and how does that answer change with audience?
Standalone/runnable via Rscript or source(). Requires only ggplot2.

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
section <- function(num, title) {
  bar <- paste(rep("=", 78), collapse = "")
  cat("\n", bar, "\n", sep = "")
  cat(sprintf("%2s. %s\n", num, title))
  cat(bar, "\n", sep = "")
}
```

## 36. WHY BAR-OF-MEANS CAN MISLEAD WITH SMALL SAMPLES (n=3 PER SCHOOL)

```{r section-36}
section(36, "WHY BAR-OF-MEANS CAN MISLEAD WITH SMALL SAMPLES (n=3 PER SCHOOL)")
students_df <- data.frame(
  Student_ID = paste0("STU-", 101:106),
  School     = rep(c("Lincoln High", "Washington High"), each = 3),
  Attendance = c(0.96, 0.82, 0.99, 0.74, 0.91, 0.88),
  Math_Grade = c(88, 54, 95, 62, 71, NA)
)

school_means <- aggregate(Math_Grade ~ School, data = students_df, FUN = mean, na.rm = TRUE)

print(
  ggplot(school_means, aes(x = School, y = Math_Grade, fill = School)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = round(Math_Grade, 1)), vjust = -0.5) +
    labs(title = "Average Math Grade by School (n=3 per school)", y = "Mean Math Grade") +
    theme_minimal() + theme(legend.position = "none")
)

cat("This bar chart LOOKS like a confident, settled comparison. But each bar\n")
cat("is the average of just 2-3 students -- with only 5 non-missing scores\n")
cat("total, that gap could easily be noise. A bar-of-means chart has no\n")
cat("visual mechanism to show you that.\n\n")

print(
  ggplot(students_df, aes(x = School, y = Math_Grade, color = School)) +
    geom_jitter(width = 0.05, size = 3, na.rm = TRUE) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black", na.rm = TRUE) +
    labs(title = "Same Data -- Every Point Shown (n honestly visible)",
         subtitle = "Black diamond = mean. Now the sample size is impossible to miss.",
         y = "Math Grade") +
    theme_minimal() + theme(legend.position = "none")
)

cat("With n this small, showing every jittered point (or a boxplot, once n\n")
cat("is large enough for one to be meaningful) is the more honest chart.\n")
cat("Bar-of-means is best reserved for when n per group is large enough that\n")
cat("the mean itself is a stable, trustworthy summary -- which is exactly\n")
cat("the situation in Section 37 below, at n=700.\n")
```

## 37. THE RIGHT COMPANION CHART FOR A TWO-WAY ANOVA: INTERACTION PLOT

```{r section-37}
section(37, "THE RIGHT COMPANION CHART FOR A TWO-WAY ANOVA: INTERACTION PLOT")
set.seed(2026)
total_students <- 700
campus_names <- c("Lincoln High", "Washington High", "Jefferson Middle", "Kennedy Elementary",
                   "Madison Academy", "Roosevelt Tech", "Garfield Charter")
races_pool <- c("Asian/Pas. Islander", "Hispanic/Latino", "Black/African Amer.", "White/Caucasian")

students_v2 <- data.frame(
  School         = sample(campus_names, total_students, replace = TRUE),
  Race_Ethnicity = sample(races_pool, total_students, replace = TRUE, prob = c(0.15, 0.35, 0.25, 0.25)),
  Math_Grade     = round(pmin(pmax(rnorm(total_students, 76, 12), 45), 100), 0)
)
students_v2$Math_Grade[sample(1:total_students, size = total_students * 0.05)] <- NA

# Collapse to the 3 largest schools for a readable interaction plot --
# all 7 schools x 4 subgroups would be 28 crossing lines, unreadable.
top3_schools <- names(sort(table(students_v2$School), decreasing = TRUE))[1:3]
plot_subset  <- subset(students_v2, School %in% top3_schools & !is.na(Math_Grade))

interaction.plot(
  x.factor = plot_subset$Race_Ethnicity,
  trace.factor = plot_subset$School,
  response = plot_subset$Math_Grade,
  fun = mean, type = "b", pch = 19, col = c("steelblue", "darkred", "darkgreen"),
  xlab = "Race/Ethnicity", ylab = "Mean Math Grade", trace.label = "School",
  main = "Interaction Plot: School x Race/Ethnicity on Math Grade"
)

cat("This is the chart a two-way ANOVA is actually FOR. Roughly parallel\n")
cat("lines mean the effect of Race/Ethnicity is consistent across schools --\n")
cat("no real interaction, matching the Type II test from Part 2 (which\n")
cat("assumes exactly that). Crossing or fanning-out lines are the\n")
cat("interaction the omnibus ANOVA p-value was testing for -- visible\n")
cat("directly here, instead of read off a table of F-statistics.\n")
```

## 38. A RELATIONSHIP CHECK MISSING FROM THE ORIGINAL SCRIPTS

```{r section-38}
section(38, "A RELATIONSHIP CHECK MISSING FROM THE ORIGINAL SCRIPTS")
set.seed(21)
attendance_grade_df <- data.frame(Attendance = pmin(pmax(rnorm(300, 0.88, 0.08), 0.5), 1))
attendance_grade_df$Math_Grade <- pmin(pmax(
  40 + 55 * attendance_grade_df$Attendance + rnorm(300, sd = 8), 0), 100)

print(
  ggplot(attendance_grade_df, aes(x = Attendance, y = Math_Grade)) +
    geom_point(alpha = 0.4, color = "steelblue") +
    geom_smooth(method = "lm", color = "darkred") +
    labs(title = "Attendance vs Math Grade -- Scatter Plot",
         x = "Attendance Rate", y = "Math Grade") +
    theme_minimal()
)

cat("Neither school_district.R nor school_district_v2.R ever plots Attendance\n")
cat("against Math_Grade directly -- yet 'does attendance predict grades' is\n")
cat("probably the single most policy-relevant question sitting in this data.\n")
cat("Per Part 3, Section 31: two numeric variables, relationship question ->\n")
cat("scatter plot is the chart built for exactly this.\n")
```

## 39. SAME UNDERLYING DATA, THREE AUDIENCES

```{r section-39}
section(39, "SAME UNDERLYING DATA, THREE AUDIENCES")
board_data <- subset(students_v2, School %in% top3_schools)
board_summary <- aggregate(Math_Grade ~ School, data = board_data, FUN = mean, na.rm = TRUE)

# ---- Version 1: School board / public-facing --------------------------------
print(
  ggplot(board_summary, aes(x = reorder(School, Math_Grade), y = Math_Grade)) +
    geom_col(fill = "#2C5F8A", width = 0.55) +
    geom_text(aes(label = round(Math_Grade, 0)), hjust = -0.3, size = 5) +
    coord_flip() +
    labs(title = "Average Math Score by Campus", x = NULL, y = NULL) +
    theme_minimal(base_size = 15) +
    theme(axis.text.x = element_blank(), panel.grid = element_blank())
)
cat("BOARD / PUBLIC VERSION: one message, one chart. No error bars, no\n")
cat("p-values, no jargon -- a board member has seconds, not minutes, and\n")
cat("doesn't need (or want) to interpret a boxplot's whiskers.\n\n")

# ---- Version 2: Internal analyst / statistics team ---------------------------
print(
  ggplot(subset(students_v2, School %in% top3_schools & !is.na(Math_Grade)),
         aes(x = School, y = Math_Grade, fill = School)) +
    geom_boxplot(alpha = 0.6, outlier.color = "red") +
    geom_jitter(width = 0.15, alpha = 0.15, size = 0.6) +
    labs(title = "Math Grade Distribution by Campus (n visible, outliers flagged)",
         subtitle = "Companion to the two-way ANOVA / Type II SS results from Part 2",
         y = "Math Grade") +
    theme_minimal() + theme(legend.position = "none")
)
cat("ANALYST VERSION: full distribution, sample size visible via jitter,\n")
cat("outliers flagged -- meant to sit next to the ANOVA table and diagnostic\n")
cat("plots from Part 2, for a reader who will ask about variance and n.\n\n")

# ---- Version 3: Equity / subgroup review --------------------------------------
students_v2$Passed <- as.integer(students_v2$Math_Grade >= 70)
equity_df <- aggregate(Passed ~ School + Race_Ethnicity,
                        data = subset(students_v2, School %in% top3_schools & !is.na(Math_Grade)),
                        FUN = mean)
names(equity_df)[3] <- "Pass_Rate"

print(
  ggplot(equity_df, aes(x = School, y = Pass_Rate, fill = Race_Ethnicity)) +
    geom_col(position = "dodge") +               # dodge, NOT fill -- see note below
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    labs(title = "Passing Rate (Math Grade >= 70) by Campus and Subgroup",
         subtitle = "Equity view: are pass rates consistent across subgroups within each campus?",
         y = "Share Passing") +
    theme_minimal()
)
cat("EQUITY VERSION: reframes the same underlying data around subgroup PASS\n")
cat("RATE rather than an overall mean -- the right chart changes because the\n")
cat("QUESTION changes, even though the source data didn't.\n\n")
cat("NOTE: this uses position='dodge', not 'fill' like Section 33's headcount\n")
cat("chart. Headcounts are genuinely parts of a whole (they sum to a\n")
cat("department total), so 100%-stacking them was correct. A pass RATE per\n")
cat("subgroup is its own independent proportion -- it is NOT a share of some\n")
cat("larger total, so stacking it to 100% would produce a number with no\n")
cat("real meaning. Confusing 'a rate' with 'a share of a whole' is a common\n")
cat("and easy mistake once you've gotten comfortable with 100%-stacked bars.\n")

cat("\n", paste(rep("=", 78), collapse = ""), "\n", sep = "")
cat("PART 4 COMPLETE.\n")
```

